---
title: "Differential expression & genotype-expression correlation of PICALM in AD, nonAD and all"
author: "XSun"
date: "2024-06-28"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

```{r warning=F, message=F}
library(ggplot2)
library(dplyr)
```

# Introduction

We address the one of the comment from the reviewer here. 

1.postmortem brain microglia AD vs. nonAD differential expression analysis of PICALM

2. genotype-expression correlation of PICALM. 

```{r warning=F, message=F}

df <- readxl::read_excel("/project2/xinhe/xsun/psych_analysis/2.review/PICALM_CPM_df_2_plot_AD_control_genotype_NG2022.xlsx",sheet=3)
colnames(df) <- c("sample","Diag","Genotype","eQTL_diagnosis","PICALM_expression","average_expr_group")

df <- df[,1:5]

df <- df %>%
  mutate(G_dosage = case_when(
    Genotype == "AA" ~ 0,
    Genotype == "AG" ~ 1,
    Genotype == "GG" ~ 2,
    TRUE ~ NA_real_
  ))

DT::datatable(df,caption = htmltools::tags$caption( style = 'caption-side: left; text-align: left; color:black;  font-size:150% ;','Individual level PICALM expression'),options = list(pageLength = 5) )
```


# Differential expression analysis of PICALM

1. The dataset was divided into 2 groups, according to the `eQTL_diagnosis` column:

- Alzheimer Disease (AD) samples
- Non-Alzheimer Disease (nonAD) samples

The black triangle represents the mean value of expressions in each group

```{r warning=F, message=F}
mean_values <- aggregate(PICALM_expression ~ eQTL_diagnosis, data = df, FUN = mean)

# Plot with mean values added
ggplot(df, aes(x = eQTL_diagnosis, y = PICALM_expression, color = eQTL_diagnosis)) +
  geom_boxplot() +
  geom_jitter(width = 0.2) +
  geom_point(data = mean_values, aes(x = eQTL_diagnosis, y = PICALM_expression), 
             color = "black", size = 3, shape = 17) +  # Add mean points
  ggtitle("PICALM Expression by Group") +
  xlab("Genotype") +
  ylab("PICALM Expression") +
  theme_minimal()
```


2. Since the expression in AD samples is normally distributed

```{r warning=F, message=F}
hist(df$PICALM_expression[df$eQTL_diagnosis =="AD"], main = "Distribution of PICALM_expression in AD samples")
```

the significance can be assessed by 

## two sample t-test

```{r warning=F, message=F}
t_test_result <- t.test(PICALM_expression ~ eQTL_diagnosis, data = df)
print(t_test_result)

ggplot(df, aes(x = eQTL_diagnosis, y = PICALM_expression, color = eQTL_diagnosis)) +
  geom_boxplot() +
  geom_jitter(width = 0.2) +
  ggtitle("PICALM Expression by Group") +
  xlab("Genotype") +
  ylab("PICALM Expression") +
  theme_minimal()


```

## ANNOVA

```{r warning=F, message=F}
anova_result <- aov(PICALM_expression ~ eQTL_diagnosis, data = df)
summary(anova_result)
```

## Linear regression

we tested PICALM_expression ~ eQTL_diagnosis

```{r warning=F, message=F}
model <- lm(PICALM_expression ~ eQTL_diagnosis, data = df)
summary(model)
```


3. Since mean value testing didn't perform well, we used other testing stats to test the median values

## Mann-Whitney U Test (Wilcoxon Rank-Sum Test)

The Mann-Whitney U test compares the median values of two independent groups. It assesses whether values in one group tend to be larger or smaller than values in another group.

```{r warning=F, message=F}
wilcox.test(PICALM_expression ~ eQTL_diagnosis, data = df)
```


# Genotype-expression correlation of PICALM

1. The dataset was divided into three groups, according to the `eQTL_diagnosis` column:

- All samples
- Alzheimer Disease (AD) samples
- Non-Alzheimer Disease (nonAD) samples

2. Correlation Analysis:

For each group (all samples, AD, and nonAD), the correlation between PICALM expression and genotype was calculated using

- Linear regression: Expr ~ genotype, we used G dosage as genotype

```{r warning=F, message=F}

df_ad <- df %>% dplyr::filter(eQTL_diagnosis == "AD")
df_nonad <- df %>% dplyr::filter(eQTL_diagnosis != "AD")

print("The group for all samples")
model_all <- lm(PICALM_expression ~ G_dosage, data = df)
summary(model_all)
sprintf("the p-value is %s", round(summary(model_all)$coefficients[2,4], digits = 4))

print("The group for AD samples")
model_ad <- lm(PICALM_expression ~ G_dosage, data = df_ad)
summary(model_ad)
sprintf("the p-value is %s", round(summary(model_ad)$coefficients[2,4], digits = 4))

print("The group for non-AD samples")
model_nonad <- lm(PICALM_expression ~ G_dosage, data = df_nonad)
summary(model_nonad)
sprintf("the p-value is %s", round(summary(model_nonad)$coefficients[2,4], digits = 4))

# Convert G_dosage back to factor for plotting
df <- df %>% mutate(Genotype = factor(G_dosage, levels = c(0, 1, 2), labels = c("AA", "AG", "GG")))
df_ad <- df_ad %>% mutate(Genotype = factor(G_dosage, levels = c(0, 1, 2), labels = c("AA", "AG", "GG")))
df_nonad <- df_nonad %>% mutate(Genotype = factor(G_dosage, levels = c(0, 1, 2), labels = c("AA", "AG", "GG")))

# Plot for all samples
ggplot(df, aes(x = Genotype, y = PICALM_expression, color = Genotype)) +
  geom_boxplot() +
  geom_jitter(width = 0.2) +
  ggtitle("PICALM Expression by Genotype (All Samples)") +
  xlab("Genotype") +
  ylab("PICALM Expression")+
  theme_minimal()

# Plot for AD samples
ggplot(df_ad, aes(x = Genotype, y = PICALM_expression, color = Genotype)) +
  geom_boxplot() +
  geom_jitter(width = 0.2) +
  ggtitle("PICALM Expression by Genotype (AD Samples)") +
  xlab("Genotype") +
  ylab("PICALM Expression")+
  theme_minimal()

# Plot for non-AD samples
ggplot(df_nonad, aes(x = Genotype, y = PICALM_expression, color = Genotype)) +
  geom_boxplot() +
  geom_jitter(width = 0.2) +
  ggtitle("PICALM Expression by Genotype (Non-AD Samples)") +
  xlab("Genotype") +
  ylab("PICALM Expression") +
  theme_minimal()

```




